Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-02 documents/ms.tex
minor edits to abstract/intro

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a 
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing effective
inference, such as having highly degenerate or high-dimensional parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
Using this highly flexible model, 
we are able to sample from the posterior Probability Density Function
(PDF) of the periodic parameter, marginalising over the other kernel hyperparameters
using a Markov Chain Monte Carlo approach.
To test this method's effectiveness, we infer rotation periods from 
\naigrain\ simulated light curves, demonstrating that our results 
    are more accurate than both a sine-fitting periodogram and an 
    AutoCorrelation Function (ACF) method.
We also demonstrate that it works effectively on real data, by 
inferring rotation periods for \nkois\ \Kepler\ stars.
Because this method delivers posteriors PDFs, 
it will enable other studies involving stellar rotation, 
such as inferring stellar
ages or stellar obliquities in exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

The light curves of spotted, rotating stars are often non-sinusoidal and
Quasi-Periodic (QP).
These stars vary in brightness due to active regions on their surfaces which
rotate in and out of view.
Complicated surface spot patterns produce the non-sinusoidal quality,
and the finite lifetimes of these active regions and the presence of 
differential rotation on the stellar surface produce quasi-periodicity.
A strictly periodic sinusoid therefore does not well describe 
rotationally modulated stellar light curves.
A physically realistic model of the stellar surface 
would perfectly capture the complexity of shapes
within stellar light curves as well as the quasi-periodic nature, allowing for
extremely precise probabilistic period recovery when conditioned on the data.
However, such physical models require many free parameters in order to
accurately represent a stellar surface and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that many stars likely have hundreds of spots, the number
of free parameters quickly becomes unwieldy, especially to explore
their posterior PDFs with MCMC.
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results; however, such 
relatively inflexible models sacrifice precision.

The standard methods used for measuring rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep{Mcquillan2013} and wavelet transforms \citep{Garcia2014}.
The precision of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram, 
and a choice of mother wavelet, assumed to desribe the data
over a range of transpositions \citep[see, \eg][]{Carter2010}, 
for the wavelet method.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, as long as it is approximately periodic the ACF will 
display a peak at the rotation period, no matter the signal shape.
A drawback of the ACF method, however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for the future large-scale survey instrument
\LSST.
An ACF is also an operation performed on the data, not a generative model of
the data, and so not inherently probabilistic.
Therefore, there is no justified way to estimate the uncertainty on an ACF
rotation period measurement.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

In this work, we introduce an {\it effective} model for rotationally
modulated stellar light curves
which captures the salient behaviour but is not
physically motivated---although some parameters may indeed be 
{\it interpreted} as physical ones.
An ideal effective model should have a small number of non-degenerate parameters 
and be flexible enough to perfectly capture non-sinusoidal, QP behaviour.
A Gaussian process (GP) model fulfills these requirements. We use this GP 
model as the generative model at the backbone of a method to probabilistically 
infer accurate and precise stellar rotation periods.  This enables us to 
estimate the posterior PDF of the rotation period, thereby producing a 
justified estimate of its uncertainty while simultaneously marginalizing over
parameters which model correlated noise.

GPs are commonly used in the machine learning community and increasingly
in other scientific fields, for example biology and geophysics.
More recently, GPs have been used in the astronomy literature \citep[see
\eg][]{Gibson2012, Haywood2014, Haywood2015, Evans2015, Rajpaul2015,
Rajpaul2016, Aigrain2016}.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a dataset is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, a GP can describe that dataset. 
An in-depth introduction to Gaussian processes is provided by
\citet{Rasmussen2005}.

GP models parameterise the covariance between data points by means of a kernel
function that defines the form of the covariance matrix.
For example, take the time-series in Figure \ref{fig:GP_example}: 
the \kepler\ light curve of KIC \kepexample, a relatively active star that
rotates once every $\sim$ 4 days, with stochastic variability typical of 
\kepler\ FGK stars.
Clearly, data points in this light curve are correlated.
Points close together in time are tightly correlated and points more
widely separated are loosely correlated.
A GP models the way in which this correlation varies with the separation between 
data points; that is, it models not the data directly but rather the {\it 
covariance structure}.
This gives GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas a much more complicated one may be
required to model the time series itself.
Figure \ref{fig:GP_example} demonstrates how the light curve of KIC \kepexample\ 
may be modelled with a GP.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions can describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE), defined as follows, 
produces an adequate fit to the KIC \kepexample\ light curve:
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay, and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple, with
just two parameters, $A$ and $l$.
If $l$ is large, two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable.
It is therefore possible to model a data set and its derivatives
simultaneously.
% \citep[see \eg][]{Rajpaul2015}

However, The SE kernel function is not a good model of the covariance 
in stellar light curves, nor a {\it useful} one for the problem of 
rotation period inference because it does not capture periodic behaviour.
In order to infer rotation periods it is thus necessary to use a periodic kernel
function.  For this reason, we use the `Quasi-Periodic' kernel, defined as
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} -
\Gamma^2 \sin^2(\frac{2\pi}{P}) \right).
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very small, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is large, points separated by one period are tightly
correlated and points separated by slightly more or less are still highly
correlated although less so.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while also allowing 
points separated by half a period to be weakly correlated.
We also use an extra parameter $\sigma$ which is an additional white noise
term added to the diagonal elements of the covariance matrix.
It is the fraction by which the observational uncertainties have been
underestimated --- if the uncertainties reported on the data are too small,
this parameter will be non-zero.
In practice however, this parameter should always be non-zero when performing
GP regression.
This is because the covariance matrix must be positive definite; however,
matrix inversion performed using most solvers is approximate, not exact,
so slight deviations from positive definiteness can arise.
Including a small amount of extra variance in the model allows enough
flexibility that the matrix inversion algorithms do not run into numerical
issues.  We use this QP kernel function to produce the GP model 
that fits the \Kepler\ light curve in in Figure \ref{fig:GP_example}.

To infer a stellar rotation period from a light curve, we fit this QP-kernel
GP model to the data.  As with any model-fitting exercise, the likelihood
of the model could be maximized to find the maximum-likelihood value for $P$.
In this study, however, we explore the full posterior PDFs using MCMC.  While
this approach comes at a computational cost, such posterior exploration 
importantly provides a justified estimate of the uncertainty on rotation period.

>>>>

\begin{abstract}
The light curves of spotted, rotating stars often vary in a non-sinusoidal and
quasi-periodic fashion---spots move on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a 
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing effective
inference, such as highly degenerate or high-dimensional parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
Using this highly flexible model, 
we are able to sample from the posterior Probability Density Function
(PDF) of the periodic parameter, marginalising over the other kernel hyperparameters
using a Markov Chain Monte Carlo approach.
To test this method's effectiveness, we infer rotation periods from 
\naigrain\ simulated stellar light curves, demonstrating that our results 
    are more accurate than both a sine-fitting periodogram and an 
    AutoCorrelation Function (ACF) method.
We also demonstrate that it works effectively on real data, by 
inferring rotation periods for \nkois\ \Kepler\ stars.
Because this method delivers posteriors PDFs, 
it will enable other studies involving stellar rotation, 
such as inferring stellar
ages or stellar obliquities in exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

The brightness of a spotted, rotating star often varies in a non-sinusoidal and
Quasi-Periodic (QP) manner, due to active regions on its surface which
rotate in and out of view.
Complicated surface spot patterns produce non-sinusoidal variations,
and the finite lifetimes of these active regions and
differential rotation on the stellar surface produce quasi-periodicity.
A strictly periodic sinusoid therefore does not well describe 
rotationally modulated stellar light curves.
A physically realistic model of the stellar surface 
would ideally perfectly capture the complexity of shapes
within stellar light curves as well as the quasi-periodic nature, allowing for
extremely precise probabilistic period recovery when conditioned on the data.
However, such physical models require many free parameters in order to
accurately represent a stellar surface, and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that stars may have hundreds of spots, the number
of free parameters in such a model quickly becomes unwieldy, especially to explore
its posterior probability distribution function (PDF).
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results; however, such 
relatively inflexible models sacrifice precision.

Standard methods to measure rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep{Mcquillan2013} and wavelet transforms \citep{Garcia2014}.
The precisions of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram, 
and a choice of mother wavelet, assumed to desribe the data
over a range of transpositions \citep[see, \eg][]{Carter2010}, 
for the wavelet method.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, as long as the signal is approximately periodic the ACF will 
display a peak at the rotation period, no matter its precise shape.
A drawback of the ACF method, however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for most ground-based measurements, such 
as the future Large Synoptic Survey Telescope (\LSST).
An ACF is also an operation performed on the data rather than a generative 
model of the data, and so is not inherently probabilistic.
Therefore, there is no justified way to estimate the uncertainty on an ACF
rotation period measurement.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

In this work, we introduce an {\it effective} model for rotationally
modulated stellar light curves
which captures the salient behaviour but is not
physically motivated---although some parameters may indeed be 
{\it interpreted} as physical ones.
An ideal effective model should have a small number of non-degenerate parameters 
and be flexible enough to perfectly capture non-sinusoidal and QP behaviour.
A Gaussian process (GP) model fulfills these requirements. We thus use a GP 
as the generative model at the core of a method to probabilistically 
infer accurate and precise stellar rotation periods.  This enables us to 
estimate the posterior PDF of the rotation period, thereby producing a 
justified estimate of its uncertainty while simultaneously marginalizing over
parameters which model correlated noise.

GPs are commonly used in the machine learning community and increasingly
in other scientific fields such as biology and geophysics.
More recently, GPs have been used in the astronomy literature \citep[see
\eg][]{Gibson2012, Haywood2014, Haywood2015, Evans2015, Rajpaul2015,
Rajpaul2016, Aigrain2016}.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a dataset is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, a GP can describe that dataset. 
An in-depth introduction to Gaussian processes is provided by
\citet{Rasmussen2005}.

GP models parameterise the covariance between data points by means of a kernel
function that defines the form of the covariance matrix.
As a qualitatively demonstration, we present the time-series 
in Figure \ref{fig:GP_example}: 
the \kepler\ light curve of KIC \kepexample, a relatively active star that
rotates once every $\sim$ 4 days, with stochastic variability typical of 
\kepler\ FGK stars.
Clearly, data points in this light curve are correlated.
Points close together in time are tightly correlated, and points more
widely separated are loosely correlated.
A GP models the way in which this correlation varies with the separation between 
data points; that is, it models not the data directly but rather the {\it 
covariance structure}.
This lends GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas modelling the time series itself
might require much more complexity.
Figure \ref{fig:GP_example} demonstrates how a GP model fits the light curve of 
KIC \kepexample.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions can describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE), defined as follows, adequately fits the 
KIC \kepexample\ light curve:
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay, and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple, with
just two parameters, $A$ and $l$.
If $l$ is large, two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable, making it possible to model a data set 
and its derivatives simultaneously.
% \citep[see \eg][]{Rajpaul2015}

However, The SE kernel function does not well describe the covariance 
in stellar light curves, nor is it {\it useful} for the problem of 
rotation period inference because it does not capture periodic behaviour.
Inferring rotation periods thus requires a periodic kernel
function.  For this reason, we use the `Quasi-Periodic' kernel, defined as
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} -
\Gamma^2 \sin^2(\frac{2\pi}{P}) \right).
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star, and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very small, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is large, points separated by one period are tightly
correlated, and points separated by slightly more or less are still highly
correlated, although less so.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while also allowing 
points separated by half a period to be weakly correlated.
We also use an extra parameter $\sigma$ which is an additional white noise
term added to the diagonal elements of the covariance matrix.
This represents the fraction by which the observational uncertainties have been
underestimated --- if the uncertainties reported on the data are too small,
it will be non-zero.
In practice, this parameter should always be non-zero when performing
GP regression, because the covariance matrix must be positive definite and  
slight deviations from positive definiteness can arise due to most 
solvers using approximate, rather than exact, matrix inversion.
Including a small amount of extra variance in the model ensures enough
flexibility for matrix inversion algorithms to not run into such numerical
issues.  We use this QP kernel function to produce the GP model 
that fits the \Kepler\ light curve in in Figure \ref{fig:GP_example}.

To infer a stellar rotation period $P$ from a light curve, we fit this QP-kernel
GP model to the data.  As with any model-fitting exercise, the likelihood
of the model could be maximized to find the maximum-likelihood value for $P$.
In this study, however, we explore the full posterior PDFs using a Markov
Chain Monte Carlo (MCMC) procedure.  While
this approach comes at a computational cost, such posterior exploration 
importantly provides a justified uncertainty estimate.
